---
title: "Hyper-heatwave predictive models"
author: "G. Brooke Anderson, Keith Oleson, Bryan Jones, Roger D. Peng"
date: "August 25, 2015"
output: html_document
---

# Overview

We have developed several predictive models to classify whether a heatwave is a hyper-heatwave, which we define as a heatwave associated with a 20% or higher increase in mortality risk in the community in which it occurs. The methods used to develop these models are fully described in a paper we are currently working on submitting. Once the paper is published, we will provide a link to the paper's web address.

This repository includes three main models, all of which use, as predictive variables, a variety of characteristics of the heatwave. These models are: 

Model           | File name
--------------  | ----------
Tree model      | `unpr.tree.rose.RData`
Bagging model   | `bag.tree.rose.RData`
Boosting model  | `boost.tree.rose.RData`

We have also included two example datasets to use to show how to use these models to explore trends in hyper-heatwaves. Those three datasets are:

Dataset                                           | File name
------------------------------------------------- | ---------
Projected heatwaves in 82 communities, 2061--2080 | `projected_heatwaves.txt`
Projected populations in 82 communities, 2061--2080 | `projected_populations.csv`
Land areas for 82 communities                      | `land_area.csv`

We explain these files further below.

Finally, for use in sensitivity analyses, we have included three health-based models developed using an alternative threshold of 19% to define hyper-heatwaves (i.e., a hyper-heatwave was defined as any heatwave associated with a 19% or higher increase in all-cause mortality risk in the community). These models are: 

Model           | File name
--------------  | ----------
Tree model      | `unpr.tree.rose19.RData`
Bagging model   | `bag.tree.rose19.RData`
Boosting model  | `boost.tree.rose19.RData`

# Loading models into R

All models are saved as R objects. Provided they are saved within the working directory for R, they can be loaded using the `load()` command. Once you load the files, these models will be available as the R objects `unpr.tree.rose`, `bag.tree.rose`, and `boost.tree.rose`:

```{r}
ls()
load("unpr.tree.rose.RData")
load("bag.tree.rose.RData")
load("boost.tree.rose.RData")
```

```{r}
ls()
```

# Example data

We have included two example data files, to show how these models can be used to explore projected trends in hyper-heatwaves. Provided the datasets are saved in R's working directory, then can be read into an R session using the following code:

```{r}
proj_hws <- read.table("projected_heatwaves.txt", header = TRUE)
proj_pops <- read.csv("projected_populations.csv", header = TRUE)
land_area <- read.csv("land_area.csv", header = TRUE)
```

### Projected heatwaves dataset

The first file, `projected_heatwaves.txt`, is a dataset of all heatwaves projected in the 82 study communities considered in our paper for the years 2061--2080. These projections were generated using the National Center for Atmospheric Research's Community Earth System's Model (CESM). There are from the second ensemble member of a large-ensemble run of CESM (CESM-LE), under the Representative Concentration Pathway 8.5 emission scenario. 

This dataset includes the following variables:

Column name          | Meaning
-------------------- | ----------------------------------
city                 | Community in which heatwave occurred (abbreviation)
start.date           | Calendar date of first day of the heatwave
end.date             | Calendar date of last day of the heatwave
mean.temp            | Average of daily mean temperature during the heatwave
max.temp             | Highest daily mean temperature during the heatwave
min.temp             | Lowest daily mean temperature during the heatwave
length               | How many days the heatwave lasted
start.doy            | Day in year that heatwave started
start.month          | Month that heatwave started
days.above.80        | Number of days in heatwave above 80F
days.above.85        | Number of days in heatwave above 85F
days.above.90        | Number of days in heatwave above 90F
days.above.95        | Number of days in heatwave above 95F
days.above.99th      | Number of days in heatwave above the community's 99th percentile temperature
days.above.99.5th    | Number of days in heatwave above the community's 99.5th percentile temperature
first.in.season      | Whether the heatwave was the first of the year to occur in its community 
mean.temp.quantile   | Quantile of the average of daily mean temperature during the heatwave
max.temp.quantile    | Quantile of the highest daily mean temperature during the heatwave
min.temp.quantile    | Quantile of the lowest daily mean temperature during the heatwave
mean.temp.1          | Average daily temperature in the community
mean.summer.temp     | Average warm season temperature in the community

Full details on the creation of this dataset are available in Oleson et al. 2015. 

### Projected populations dataset

The second file, `projected_populations.csv`, is a dataset that includes the projected populations of all 82 study communities for 2061--2080 under two different population scenarios, Shared Socioeconomic Pathways 3 and 5. 

This dataset has the following variables: 

Column name          | Meaning
-------------------- | ----------------------------------
city                 | Community (abbreviation)
county               | County name
statename            | State name
fips                 | FIPS county code
SSP3                 | Projected population under Shared Socioeconomic Pathway 3
SSP5                 | Projected population under Shared Socioeconomic Pathway 5

*Note: Since some of the study communities cover more than one county, there are multiple rows for some study communities in this dataset.* 

### Land area dataset

The third file, `land_area.csv`, provides the land area of each of the study communities. It has the following variables:

Column name          | Meaning
-------------------- | ----------------------------------
city                 | Community (abbreviation)
arealand             | Land aread of the community

# Formatting a heatwave dataset for the models

To run through the predictive models, data describing heatwaves must be in the format of a dataframe and must include all of the heatwave characteristics in `proj_hws`, plus two characteristics generated from the population data, community population and community population density. 

Here is an example of code to add the variables `pop100`, which gives population size under the SSP5 scenario, and `pop.density`, which gives the population density under the SSP5 scenario. This code uses functions from the `dplyr` function to sum populations across different counties for a community, join the population data to the land area data by community, and then calculate the population density.

```{r, message=FALSE}
library(dplyr)

proj_pops <- group_by(proj_pops, city) %>%
        summarise(pop100 = sum(SSP3)) %>%
        left_join(land_area) %>%
        mutate(pop.density = pop100 / arealand) %>%
        dplyr::select(-arealand)
head(proj_pops)
```

Now this data can be merged in with the heatwave characteristics given in `proj_hws`:

```{r}
proj_hws <- left_join(proj_hws, proj_pops, by = "city")
```

The `proj_hws` dataset now has all the variables needed to run through the health-based models: 

```{r}
colnames(proj_hws)
```

To use these health-based models for a dataset of heatwaves, it is important that all the same variables be calculated for the heatwaves and that they be saved using the same column names as those used here. This is how the health-based models will recognize the heatwave characteristics required to predict from the model.

# Using the models

Once a heatwave dataset is in the correct format, the models can be applied to the dataset to predict which heatwaves are likely to be hyper-heatwaves, based on these heatwave characterstics. Since the models generate a lot of false positives, the total number of projected hyper-heatwaves will then need to be adjusted to account for the models' false positive rates (precision). 

### Generating predictions from the models

To have access to prediction methods for the three models, it is necessary to load the following three R packages:

```{r, message = FALSE}
library(tree)
library(randomForest)
library(gbm)
```

For all three models, the `predict` function can be used to use the model to predict outcomes for a new dataset, in this case the heatwave dataset `proj_hws`:

```{r}
projs_tree <- predict(unpr.tree.rose,
                      newdata = proj_hws, 
                      type = "class")

projs_bag <- predict(bag.tree.rose,
                     newdata = proj_hws)

projs_boost <- predict(boost.tree.rose,
                       newdata = proj_hws, 
                       n.trees = 500)
```

The methods for these three models differ somewhat, and as a result, the prediction calls differ and the predictive output differs in format. For example, while the predictions from the tree and bagging models are a string of predictions for the heatwaves-- either "very" if the heatwave is predicted to be a hyper-heatwave or "other" otherwise-- the predictions from the boosting model are numeric. 

```{r}
head(projs_tree)
head(projs_bag)
head(projs_boost)
```

The outputs from the boosting model can be converted to the same format as the other predictions using the following code:

```{r}
projs_boost <- factor(ifelse(projs_boost > 0, "very", "other"))
head(projs_boost)
```

### Estimating hyper-heatwave frequency from the predictions

The following equations can be used to estimate the total number of hyper-heatwaves projected within a dataset of heatwaves: 

$$ H_{h} = H_{ph} * P + H_{pl} * F $$

where: 

- $H_{h}$ is the estimated number of hyper-heatwaves
- $H_{ph}$ is the number of heatwaves predicted by the model to be hyper-heatwaves
- $P$ is the model's estimated precision
- $H_{pl}$ is the number of heatwaves predicted by the model to be less dangerous heatwaves
- $F$ is the model's estimated false precision rate

This equation takes the total number of heatwaves predicted by the model to belong to each class and adjusts these values by estimates of the model's precision and false omission rates. 

The equation requires estimates of the precision and false omission rates. We estimated these for each of the three models using Monte Carlo cross-validation (for more details, see our paper describing the development of these models). These are our estimates for these values for the three models:

Model          | Precision   | False omission rate
-------------- | ----------- | --------------------
Tree model     | 2.6%        | 0.0%
Bagging model  | 2.6%        | 0.0%
Boosting model | 2.3%        | 0.0%

Here is an example of how to apply this equation to projections from the bagging model. First, a vector is created that has the adjustment rates (false omission rates and precision rates). 

```{r}
adjustment_rates <- c(0.0, 2.6) / 100

unadj_counts <- table(projs_bag)
unadj_counts

sum(unadj_counts * adjustment_rates)
```

Based on this calculation, our projection for this set of scenarios is that we would expect around `r round(sum(unadj_counts * adjustment_rates / 100))` hyper-heatwaves in the 82 study communities for 2061--2080. 

### Estimating hyper-heatwave frequency from the predictions

Exposure to hyper-heatwaves combines both the frequency of such heatwaves with projections of how long they will last and how many people will be living in the affected community. The following equation can be used to project hyper-heatwave exposure using the predictions from our health-based models: 

$$ E_{h} = P\sum_{i}(D_{i}N_{i}) + F\sum_{j}(D_{j}N_{j}) $$

where: 

- $E_{h}$ is the estimated exposure to hyper-heatwaves, in person-days
- $P$ is the model's estimated precision
- $i$ indexes across all heatwaves classified as hyper-heatwaves by the predictive model
- $D_{i}$ is the length of heatwave $i$ in days
- $N_{i}$ is the population of the community in which heatwave $i$ occurs
- $F$ is the model's estimated false omission rate
- $j$ indexes across all heatwaves classified as less dangerous heatwaves by the predictive model
- $D_{j}$ is the length of heatwave $j$ in days
- $N_{j}$ is the population of the community in which heatwave $j$ occurs

The following code gives an example of how exposure to hyper-heatwaves could be estimated in R based on `projs_bag`, the results from applying the predictive bagging model to the projected heatwave data.

First, it's necessary to combine the model predictions with data from the original heatwave dataset on heatwave length and population of the community in which the heatwave occurred:

```{r}
exp_projs <- data.frame(prediction = projs_bag,
                        length = proj_hws$length,
                        pop = proj_hws$pop100)
exp_projs[sample(1:nrow(exp_projs), 6), ]
```

Now this dataframe can be used to calculate the person-days of exposure for each heatwave and sum this exposure by predicted class of heatwave:

```{r}
exp_projs <- mutate(exp_projs, exposure = length * pop) %>%
        group_by(prediction) %>%
        summarise(exposure = sum(exposure))
exp_projs
```

Now the estimated exposure can be calculated by adjusting these two values for the model's estimated precision and false omission rates (saved in earlier code as `adjustment_rates`):

```{r}
sum(exp_projs$exposure * adjustment_rates)
```

This number can also be presented in units of millions of person-days per year (the example dataset covers 20 years):

```{r}
sum(exp_projs$exposure * adjustment_rates) / 10^6 / 20
```

For this ensemble member, emission scenario, and population scenario, our estimated projection of exposure to hyper-heatwaves is `r round(sum(exp_projs$exposure * adjustment_rates) / 10^6 / 20)` million person-days per year across the 82 study communities.